troubleshooting

### 2024-06-04 README.Rmd
update to man

---
output: github_document
---

@@

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```

@@

<!-- badges: start -->
[![R-CMD-check](https://github.com/baynec2/mspms/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/baynec2/mspms/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

@@

You can install the released version of mspms from github

```{r,eval = FALSE}
devtools::install_github("baynec2/mspms")
```

@@

There are 4 different types of functions in this package.  Those that are involved in:

@@

3. Statistics. These methods allow the user to perform basic statistics on the normalyzed/processed data.

4. Data visualization. These functions allow the user to visualize the data in a number of useful way.

**Making mspms generically useful**.   
1. *prepare_peaks()*:Takes two input files from PEAKS and combines them. 
2. *prepare_pd()*: prepares exported files from proteome discoverer.  
3. *extract_design_matrix_pd()*: Extracts the design matrix from the proteome discoverer file.  
4. *prepare_spectronaut()*: prepares exported files from spectronaut - note in development, not yet implemented.  
5. *calculate_all_cleavages()*: Calculates all possible cleavages for peptide library sequences.

**Data Proessing/ Normalization**.  
1. *normalyze()*: Normalizes values.     
2. *handle_outliers()*: Looks for outliers across replicates. Removes them.       
3. *impute()*: Imputes data for missing values (not including NAs introduced by handle_outliers()).    
4. *join_with_library()*: Joins the normalyzed data with the peptide library sequences.  
5. *add_cleavages()*: Figures out the locations of the detected clevages within the library of peptide sequences used. Is it cleaved at the N or C terminus, or both?  Uses *cterm_cleavage()* and *nterm_cleavage()* to do this.    
6. *polish()*: combines the cleavage info into one column. Discards peptides that were cleaved on both sides or not at all.  
7. *prepare_for_stats()*: Reshapes the data into a long format and appends the data in the design matrix to make statistical testing easy to perform.

**Statistics**.   
1. *log2fc_t_test()*: Calculates the fold change and the p/q value across experimental conditions using T-tests.      
2. *mspms_anova()*: Conducts ANOVA tests to determine if there are any peptides that are significantly different across all time points.

**Data Visualization**.   
1. *plot_heatmap()*: Conducts hierarchical clustering analysis and plots an interactive heatmap to visualize overall patterns in the data.    
2. *plot_pca()*: PCA analysis to visualize the data in a 2D space.  
3. *plot_time_course()*: Plot peptides over time by condition.  
4. *plot_cleavage_motif()*: Visualize the sequence specificity of the cleavage sites. Designed to be equivalent to what is implemented in IceLogo.

@@

### Loading the files ###
lfq_filename = "tests/testdata/protein-peptides-lfq.csv"
#file "protein-peptides.csv" exported from PEAKS identification
id_filename = "tests/testdata/protein-peptides-id.csv"

# Prepare the data for normalyzer analysis
peaks_prepared_data = prepare_peaks(lfq_filename,id_filename)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(peaks_prepared_data,overwrite = TRUE)
```

@@

```{r}
prepared_proteome_discoverer = prepare_pd("tests/testdata/proteome_discoverer_output.xlsx")
```

We can also extract a design matrix from the naming scheme used by proteome discoverer as follows.

```{r}
 design_matrix = extract_design_matrix_pd("tests/testdata/proteome_discoverer_output.xlsx")
```

@@

```{r}
all_peptide_sequences = mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
                        n_AA_after_cleavage=4)
head(all_peptide_sequences)

@@

```{r}
all_peptide_sequences = mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
                        n_AA_after_cleavage=5)
head(all_peptide_sequences)
```

@@

Before we normalyze data, we need to know what samples are in what groups. We can do that by defining a design matrix. 
```{r}
design_matrix = readr::read_csv("tests/testdata/design_matrix.csv")

head(design_matrix)

```
### Normalyzing data

@@

We can normalyze the data as follows:  
```{r}
normalyzed_data = normalyze(peaks_prepared_data,design_matrix)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(normalyzed_data,overwrite = TRUE)
```

@@

Here we use a dixon test from the outliers package to detect outliers from each of our replicates. We need to know what samples are part of which groups, so we need to specify the design matrix here too. Make sure that the column header names are "sample" and "group" just like before.

```{r}
design_matrix = readr::read_csv("tests/testdata/design_matrix.csv")
outliers = handle_outliers(normalyzed_data,design_matrix)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(outliers,overwrite = TRUE)
```

@@

```{r}
imputed = impute(outliers)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(imputed,overwrite = TRUE)
```

@@

Next we need to join everything with the sequences of the peptide library.  
```{r}
joined_with_library = join_with_library(imputed)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(joined_with_library,overwrite = TRUE)
```

@@

Sequences are presented as the user specified number of amino acids on both sides of a cleavage. The default is 4, but there is interest in looking at motifs that are farther away from the cut site. Note that X indicates that there was nothing on that side in the library because it was cleaved close to the edge.

```{r}
cleavage_added_data = add_cleavages(joined_with_library,n_residues = 4)

head(cleavage_added_data)

# saving to data folder in package for tests. Ignore when using on your own data
usethis::use_data(cleavage_added_data,overwrite = TRUE)
```

@@

Someimes there is a need to polish the data a bit for downstream analysis. This function does that by removing combining the cterm and nterm cleavage information into one column while removing any rows that don't have any cleavage information or have cleavage information on the cterm and nterm.

```{r}
polished_data = polish(cleavage_added_data)
```

@@

```{r}
prepared_for_stats = prepare_for_stats(cleavage_added_data,design_matrix)

usethis::use_data(prepared_for_stats,overwrite = TRUE)
```

Noe we can conduct the statistics.

@@

Ttests are performed within each condition and compared to time 0.

@@

```{r}
# Perform T test
t_test_stats = mspms::mspms_t_tests(prepared_for_stats)
```

@@

We can also calculate the log 2 fold change. The comparisons here are the same as for the T tests.  
```{r}
# calclulate log2fc
log2fc = mspms::mspms_log2fc(prepared_for_stats)

@@

Sometimes (such as when you want to make volcano plots) it is useful to look at the log2fc and the t test statistics at the same time. This function does that.

```{r}
log2fc_t_test = mspms::log2fc_t_test(prepared_for_stats)

@@

Now that we have this data we can visualize these results easily using the ggplot2 package as follows.

```{r}
library(ggplot2)

p1 = log2fc_t_test %>% 
  ggplot(aes(x = log2fc,y = -log10(p.adj)))+
  geom_point(size =0.5)+
  geom_hline(yintercept = -log10(0.05),linetype = "dashed",color = "red")+
  geom_vline(xintercept = 3, linetype = "dashed",color = "red")+
  geom_vline(xintercept = -3, linetype = "dashed",color = "red")+
  theme_minimal()+
  labs(x = "Log2 Fold Change",y = "-log10(p value)")+
  facet_wrap(~comparison,scales = "free")

@@

We also might want to perfom an anova. Here, we have it set up to show the effect of time within each condition.

@@

```{r}
# Doing ANOVA
anova_stats = mspms::mspms_anova(prepared_for_stats)

@@

@@

Here use an approach similar to what is implemented in ICELogo to visualize the sequence specificity of the cleavage sites. This was used as reference to build the code: https://iomics.ugent.be/icelogoserver/resources/manual.pdf.

```{r}
cleavage_seqs = mspms::prepared_for_stats %>%
  mspms::polish() %>% 
  filter(condition == "DMSO",time == 240) %>%
  pull(cleavage_seq)

background_universe = mspms::all_possible_8mers_from_228_library

mspms::plot_cleavage_motif(cleavage_seqs,background_universe)
```

### PCA

@@

```{r}
mspms::plot_pca(prepared_for_stats)
```

@@

We can also generate an interactive heatmap with the hierchical clustering results.

@@

```{r,eval = FALSE}

mspms::plot_heatmap(prepared_for_stats)

@@

This is best used in combination with the ANOVA function. Below, we will use anova to calculate the peptides where we see an effect of time (in the DMSO group), and then plot these.

@@

top10sig = anova_stats %>% 
  arrange(p.adj) %>%
  pull(Peptide) %>% 
  head(10)

p1 = prepared_for_stats %>%
  dplyr::filter(Peptide %in% top10sig) %>%
  mspms::plot_time_course()

@@

>>>>

---
output:
  github_document: default
  self_contained: default
---

@@

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)

ggplot2::theme_set(
  ggplot2::theme_minimal()
)

@@

<!-- badges: start -->
[![R-CMD-check](https://github.com/baynec2/mspms/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/baynec2/mspms/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/baynec2/mspms/branch/main/graph/badge.svg)](https://app.codecov.io/gh/baynec2/mspms?branch=main)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/mspms.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/mspms)
<!-- badges: end -->

@@

You can install the development version of mspms from github.

```{r,eval = FALSE}
devtools::install_github("baynec2/mspms",
                         ref = "develop")
```

## Quickstart

To generate a general report using your own data, run the following code. 
It requires data that has been prepared for mspms data analysis by a 
converter function, and a design matrix. For more information see subsequent 
sections.

```{r,eval = FALSE}
mspms::generate_report(
  prepared_data = mspms::peaks_prepared_data,
  design_matrix = mspms::design_matrix,
  outdir = "../Desktop/mspms_report"
)
```

The above command will generate a .html file containing a general mspms analysis.

There is much more that can be done using the mspms package- see the following 
sections for more information.

@@

There are 4 different types of functions in this package.

@@

3. Statistics. These methods allow the user to perform basic statistics on the 
normalized/processed data.

4. Data visualization. These functions allow the user to visualize the data.

**Making mspms generically useful**.   
1. *prepare_peaks()*:Takes two input files from PEAKS and combines them. 
2. *prepare_pd()*: prepares exported files from proteome discoverer.  
3. *calculate_all_cleavages()*: Calculates all possible cleavages for peptide
library sequences.

**Data Processing/ Normalization**.  
1. *normalyze()*: Normalizes values.     
2. *handle_outliers()*: Looks for outliers across replicates. Removes them.       
3. *impute()*: Imputes data for missing values (not including NAs introduced 
by handle_outliers()).  
4. *join_with_library()*: Joins the normalized data with the peptide library 
sequences.  
5. *add_cleavages()*: Figures out the locations of the detected cleavages within
the library of peptide sequences used. Is it cleaved at the N or C terminus, 
or both? Uses *cterm_cleavage()* and *nterm_cleavage()* to do this.      
6. *polish()*: combines the cleavage info into one column. Discards peptides
that were cleaved on both sides or not at all.  
7. *prepare_for_stats()*: Reshapes the data into a long format and appends the 
data in the design matrix to make statistical testing easy to perform.  
8 *mspms():* combines all of the above functions into one convenient function.

**Calculations/ Statistics**.   
1. *mspms_log2fc()*: Calculates the log2 fold change across experimental 
conditions relative to T0.  
2.*mspms_t_tests()*: Conducts T-tests to determine if there are any peptides 
that are significantly different across experimental conditions relative to time 
0.  
3. *log2fc_t_test()*: Combines results from mspms_log2fc() and mspms_t_tests() 
into one convenient tibble.   
4. *log2fct_condition()*: Calculates the log2 fold change and t tests for the 
specified condition at a user specified time.    
5.*log2fct_time()*: Calculates the log2 fold change and t tests between the 
specified times within a user specified time.   
6. *mspms_anova()*: Conducts ANOVA tests to determine if there are any peptides 
that are significantly different across all time points.   
7. *count_cleavages_per_pos():* Counts the number of cleavages at each position 
of the library. Useful for determining endoprotease vs exoprotease activity.

**Data Visualization**.   
1. *plot_heatmap()*: Conducts hierarchical clustering analysis and plots an
interactive heatmap to visualize overall patterns in the data.    
2. *plot_pca()*: PCA analysis to visualize the data in a 2D space.  
3. *plot_time_course()*: Plot peptides over time by condition.  
4. *plot_cleavages_per_pos()*: Plot the number of cleavages at each position 
in the library.Useful for determining endoprotease vs exoprotease activity.

**Icelogo**.

1. *calc_AA_count_of_motif()*: calculates the count of each amino acid at each
position in a vector of motifs.  
2. *calc_AA_prop_of_motif()*: calculates the proportion of each amino acid at 
each position from the output of calc_AA_count_of_motif().  
3. *calc_AA_motif_zscore()*: calculates z scores for each amino acid at each 
position in a vector of motifs.  
4. *calc_AA_percent_difference()*: calculates the percentage difference of 
5. *prepare_sig_p_dif()*: prepares the data for plotting percent difference 
icelogos. Removes AAs at positions that are not significantly different.  
6. *prepare_fc()*: prepares the data for plotting fold change icelogos.
Removes AAs at positions that are not significantly different.  
7. *plot_pd_icelogo()*: plots a percent difference icelogo.  
8. *plot_fc_icelogo()*: plots a fold change ice logo.  
9. *plot_icelogo()*:  wraps plot_pd_icelogo() and plot_fc_icelogo() into 
one function.  
10. *plot_all_icelogos()*: convenience function to plot all icelogos relative to 
time 0 from an experiment.    
**Reports**. 
1. *generate_report()*: Generates a simple analysis of the mspms data that 
should be generically applicable to all mspms experiments.

@@

### Loading the files ###
lfq_filename <- "tests/testdata/protein-peptides-lfq.csv"
# file "protein-peptides.csv" exported from PEAKS identification
id_filename <- "tests/testdata/protein-peptides-id.csv"

# Prepare the data for normalyzer analysis
peaks_prepared_data <- prepare_peaks(lfq_filename, id_filename)
```

@@

```{r}
prepared_proteome_discoverer <- prepare_pd("tests/testdata/proteome_discoverer_output.xlsx")
```

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 4
)
head(all_peptide_sequences)
```

@@

```{r}
all_peptide_sequences <- mspms::calculate_all_cleavages(mspms::peptide_library$library_real_sequence,
  n_AA_after_cleavage = 5
)
head(all_peptide_sequences)
```

@@

Before we normalyze data, we need to know what samples are in what groups. We can do that by defining a design matrix. 
```{r}
design_matrix <- readr::read_csv("tests/testdata/design_matrix.csv")

head(design_matrix)
```
### Normalyzing data

@@

We can normalyze the data as follows:  
```{r}
normalyzed_data <- normalyze(peaks_prepared_data, design_matrix)

@@

Here we use a Dixon test from the outliers package to detect outliers from each 
of our replicates. We need to know what samples are part of which groups, so we 
need to specify the design matrix here too. Make sure that the column header 
names are "sample" and "group" just like before.

```{r}
design_matrix <- readr::read_csv("tests/testdata/design_matrix.csv")
outliers <- handle_outliers(normalyzed_data, design_matrix)

@@

```{r}
imputed <- mspms::impute(outliers)

@@

Next we need to join everything with the sequences of the peptide library.  
```{r}
joined_with_library <- mspms::join_with_library(imputed)
```

@@

Sequences are presented as the user specified number of amino acids on 
both sides of a cleavage. The default is 4, but there is interest in looking at
motifs that are farther away from the cut site. Note that X indicates that there
was nothing at that position because it is over the end.

```{r}
cleavage_added_data <- mspms::add_cleavages(joined_with_library, n_residues = 4)

head(cleavage_added_data)
```

### Adding metadata back

We need to add information about our samples back to perform downstream 
statistics on the data. We accomplish this with the prepare for stats file.

```{r}
prepared_for_stats <- mspms::prepare_for_stats(cleavage_added_data,
                                               design_matrix)
```

@@

This function polishes the data by combining the cterm and nterm cleavage 
information into one column while removing any rows that don't have any 
cleavage information or have cleavage information on the cterm and nterm.

```{r}
polished_data <- mspms::polish(cleavage_added_data)
```

### Complete workflow

There are a lot of steps to get to the final normalized data we listed above. 
We have combined all of these steps into one function, mspms() for convenience.
This offers a nice balance between allowing the user the option to use 
individual steps and providing a quick way to get to the final data.

```{r}
mspms_data <- mspms::mspms(peaks_prepared_data, design_matrix)
```

@@

Now we can conduct the statistics:

@@

T tests are performed within each condition and compared to time 0.

@@

```{r}
# Perform T test
t_test_stats <- mspms::mspms_t_tests(mspms_data)
```

@@

We can also calculate the log 2 fold change. 
The comparisons here are the same as for the T tests.  
```{r}
# calculate log2fc
log2fc <- mspms::mspms_log2fc(mspms_data)

@@

Oftentimes (such as when you want to make volcano plots) it is most useful to
look at the log2fc and the t test statistics at the same time.

```{r}
log2fc_t_test <- mspms::log2fc_t_test(mspms_data)

@@

### log2fct_condition

You may wish to compare two conditions at a specific time point. 
This function allows this

```{r}
condition_t = mspms::log2fct_condition(mspms_data,
                                       ref_condition = "DMSO",
                                       comparison_condition = "MZB",
                                       at_time = 60)

head(condition_t)

### log2fct_time

You may wish to compare two time points within a specific condition. 
This function allows

```{r}
time_t = mspms::log2fct_time(mspms_data,
                             within_condition = "DMSO",
                             ref_time = 0,
                             comparison_time = 60)

head(time_t)
```

@@

We also might want to perform an anova. Here, we have it set up to show the 
effect of time within each condition.

@@

```{r}
# Doing ANOVA
anova_stats <- mspms::mspms_anova(mspms_data)

@@

### volcano plots

```{r}
volcano_plot <- mspms::plot_volcano(log2fc_t_test)

volcano_plot
```

@@

Here use an approach similar to what is implemented in ICELOGO to visualize the
sequence specificity of the cleavage sites. This was used as reference to build 
the code: https://iomics.ugent.be/icelogoserver/resources/manual.pdf.

Below we can see this applied to see what cleavage sequences are enriched among
the peptides significantly different between time 0 and time 60 
within DMSO.

```{r}
cleavage_seqs <- time_t %>%
  dplyr::filter(log2fc > 3, p.adj <= 0.05) %>%
  dplyr::pull(cleavage_seq)

background_universe <- mspms::all_possible_8mers_from_228_library

mspms::plot_icelogo(cleavage_seqs, background_universe)
```

We could also look at the fold change instead of the percent difference

```{r}
mspms::plot_icelogo(cleavage_seqs, background_universe, type = "fold change")
```

We also provide a convenience function for looking at all icelogos relative to 
time 0 within the experiment.

This function includes cleavages that were significant at any time point once in 
the experimental set of the ICELOGO.

```{r}
mspms::plot_all_icelogos(mspms_data)
```

### PCA.

@@

```{r}
mspms::plot_pca(mspms_data)
```

@@

We can also generate an interactive heatmap with the hierarchical clustering 
results.

@@

```{r,eval = FALSE}
mspms::plot_heatmap(mspms_data)
```

@@

This can be used in combination with the ANOVA function.
Below, we will use anova to calculate the peptides where we see an 
effect of time (in the DMSO group), and then plot these.

@@

top10sig <- anova_stats %>%
  dplyr::arrange(p.adj) %>%
  dplyr::pull(Peptide) %>%
  head(10)

p1 <- mspms_data %>%
  dplyr::filter(Peptide %in% top10sig) %>%
  mspms::plot_time_course()

@@

## Calculating the number of cleavages at each position in the library

We can also calculate the number of cleavages at each position of the library.
Here we will do this for peptides significantly different relative to time 0 for
DMSO and MZB.

```{r}
# Performing T tests
t_tests = log2fc_t_test(mspms_data) %>% 
  dplyr::filter(p.adj < 0.05,
                log2fc > 3)

count = mspms::count_cleavages_per_pos(t_tests)

p1 = mspms:::plot_cleavages_per_pos(count)
